Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-06-26 paper.tex
Fix fig 6 size

Tools for estimating geographical distribution of cycling potential have
advanced substantially in recent years. Developed by researchers and
developers in academic, public and private sectors, they have evolved
from simple area-based static models tailored to specific regions to
more complex and (in some cases) more generalizable tools. These tools
can be considered as a specific application of the concept of planning
support systems (PSS), which are designed to help planners and
decision-makers to make more informed decisions \citep{geertman2009}.
Outputs include maps with results available at several levels of
analysis with levels of availability ranging from being only available
to researchers and in static maps to open access mapping systems
available to the public.

@@

Area-based approaches were developed before network-based tools, such as
a 2007 regression-based model to estimate the percentage of cycling
trips to work in England and Wales based on socioeconomic,
transportation, and physical factors at the small area (wards) level in
2007 \citep{parkin_estimation_2007}. They found that areas with a higher
percentage of females, non-whites, car ownership, lower socioeconomic
classes, income, distance to work, population density, poor highway
conditions, hills, rainfall, and fewer off-road bicycle routes tended to
have a lower proportion of cycling to work. The approach also generated
``forecasts for potential levels of bicycle use'' with with increases in
off-road routes, improvements of highway and, conversely, increases in
car ownership and distance to work leading to decreases in cycling to
work.

Three years later, Transport for London created the Analysis of Cycling
Potential (ACP) tool based on a mode choice model that assesses the
likelihood of trips being cycled, with data from the TfL London Travel
Demand Survey as its primary input. The tool generates tables, graphs,
and heatmaps showing the potential for cycling for different trips
purposes in London to guide local cycling initiatives, such as where new
hire `docking stations' should go. The tool does not use
origin-destination data and cannot estimate cycling potential on
specific routes \citep{transport_for_london_analysis_2010}.

Other tools that generate areal data to prioritize investment include a
study that generated a {`bikeability index' for the Metro Vancouver
region \citep{winters_mapping_2013}, later renamed Bike Score and
available for over 160 US and Canadian cities \citep{winters_bike_2016}
-- its methodology is
\href{https://www.walkscore.com/bike-score-methodology.shtml}{openly
accessible}; a} `prioritization index' for Montreal, Canada
\citep{larsen_build_2013}; a `usage intensity index' based on stated
preferences in Belo Horizonte, Brazil \citep{zhang_prioritizing_2014};
and a Cycling Potential Tool (CPT) for Scotland
\citep{phillips_development_2017}. The CPT is composed of the base
environmental module --- based on eight weighted factors (population
density, hilliness, physical barriers, access to services, existing
cycling mode share, distance to work and school, and road speed) --- and
the quality of service module, which evaluates each area of interest
based on eight factors (surface condition, adjacent cyclists, comfort
factor, conflict, distance between junctions, slope, access to services,
and origin/destination).

@@

Route-based tools emerged with increasing availability of
origin-destination data, routing engines that can assign trips to
networks, and improvements in computer hardware and software needed to
generate route networks. A prominent example is the the Propensity to
Cycle Tool (PCT), first developed to estimate current and future levels
of cycling at desire line, zone, route, and route network levels for
case study cities \citep{lovelace2016}. The approach was scaled-up to
estimate estimate the potential benefits of uptake at zone and desire
line levels nationally, and launched as a publicly available web
application in 2017 \citep{lovelace2017}. Extensions of the PCT approach
have included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modeling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

{Another route-based tool is The Cycling Routing Algorithm for Network
Connectivity (CRANC) \citep{gehrke_cycling-focused_2020}. The CRANC tool
evaluates the benefits of new bicycle facilities for the accessibility
of various populations and neighbourhoods.}

@@

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE Tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
Tool, publicly available at \href{https://cruse.bike/}{cruse.bike}.}

@@

Following feedback from stakeholders, we added another, `non-everyday'
trip purpose in an extension phase: cycle tourism. As outlined in the
CRUSE Extension report {[}reference and link to be added on
publication{]}, this involved developing an inter-county spatial
interaction model to estimate the number of trips between each county
for cycle tourism, with trip attractors including campsites and major
international transport hubs.

@@

The datasets outlined above were used to generate estimates of cycling
currently (the `Baseline' scenario) and under four scenarios of cycling
potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
of these scenarios is described below, as illustrated in
Figure~\ref{fig-scenarios}.

@@

\caption{\label{fig-landing}Illustration of the landing page of the
CRUSE Tool for Ireland, showing the default `zoomed out' view of the
national map, with county level results showing the proportion of trips
cycled under the `Go Dutch' scenario (top). The default `zoomed out'
view enables users to click on a county for more detailed results. The
`zoomed in' view shows the national network view for the `Go Dutch'
scenario for the Limerick area (bottom).}

@@

The stakeholder engagement was carried out through workshops with
stakeholders in case study area, an approach that is documented with
reference to participatory development of digital tools for supporting
energy transitions \citep{hewitt2020}. We held a total of three
workshops over the course of the project and one open-ended live demo
during a 2-day event in Sligo in which we collected feedback from dozens
of people. Other than the live demo, all of the workshops took place
online.

The first two workshops took place in May 2022, during which the tool
was introduced to key staff involved in cycle network planning in
Kildare and Limerick counties. The county-specific approach allowed the
workshop to focus on specific areas and key corridors. The local
transport planners were asked to provide feedback in a semi-structured
interview approach, with questions on the evidence they currently have
access, how they may be able to use the tool, and requests for new
features. The third and final workshop took place on the
7\textsuperscript{th} December 2023 and included descriptions of recent
changes to the tool, with an aim being to get feedback on the new
recreational and cycle touring evidence, with representatives from
county-level transport planning bodies, Sport Ireland, and Waterways
Ireland.

@@

\includegraphics{rnet_types.png}

>>>>

Tools for estimating geographical distribution of cycling potential have
advanced substantially in recent years. Developed by researchers and
developers in academic, public and private sectors, they have evolved
from simple area-based static models tailored to specific regions to
more complex and (in some cases) more generalizable tools. These tools
can be considered as a specific application of the concept of planning
support systems (PSS), which are designed to help planners and
decision-makers in making better-informed decisions
\citep{geertman2009}. Outputs include maps with results available at
several levels of analysis with levels of availability ranging from
being only available to researchers and in static maps to open access
mapping systems available to the public.

@@

Area-based approaches were developed before network-based tools, such as
a 2007 regression-based model to estimate the percentage of cycling
trips to work in England and Wales based on socioeconomic,
transportation, and physical factors at the small area (wards) level in
2007 \citep{parkin_estimation_2007}. They found that areas with a higher
percentage of females, non-whites, car ownership, lower socioeconomic
classes, income, distance to work, population density, poor highway
conditions, hills, rainfall, and fewer off-road bicycle routes tended to
have a lower proportion of cycling to work. The approach also generated
``forecasts for potential levels of bicycle use'' with increases in
off-road routes, improvements of highway and, conversely, increases in
car ownership and distance to work leading to decreases in cycling to
work.

Three years later, Transport for London created the Analysis of Cycling
Potential (ACP) tool based on a mode choice model that assesses the
likelihood of trips being cycled, with data from the TfL London Travel
Demand Survey as its primary input. The ACP tool generates tables,
graphs, and heatmaps showing the potential for cycling for different
trips purposes in London to guide local cycling initiatives, such as
where new hire `docking stations' should go. However, it does not use
origin-destination data and cannot estimate cycling potential on
specific routes \citep{transport_for_london_analysis_2010}.

Other tools that generate areal data to prioritize investment include a
study that generated a {`bikeability index' for the Metro Vancouver
region \citep{winters_mapping_2013}, later renamed Bike Score and
available for over 160 US and Canadian cities \citep{winters_bike_2016};
a} `prioritization index' for Montreal, Canada
\citep{larsen_build_2013}; a `usage intensity index' based on stated
preferences in Belo Horizonte, Brazil \citep{zhang_prioritizing_2014};
and a Cycling Potential Tool (CPT) for Scotland
\citep{phillips_development_2017}. The CPT is composed of the base
environmental module --- based on eight weighted factors (population
density, hilliness, physical barriers, access to services, existing
cycling mode share, distance to work and school, and road speed) --- and
the quality of service module, which evaluates each area of interest
based on eight factors (surface condition, adjacent cyclists, comfort
factor, conflict, distance between junctions, slope, access to services,
and origin/destination).

@@

Route-based tools emerged with increasing availability of
origin-destination data, routing engines that can assign trips to
networks, and improvements in computer hardware and software needed to
generate route networks. A prominent example is the the Propensity to
Cycle Tool (PCT), first developed to estimate current and future levels
of cycling at desire line, zone, route, and route network levels for
case study cities \citep{lovelace2016}. The approach was scaled-up to
estimate the potential benefits of uptake at zone and desire line levels
nationally, and launched as a publicly available web application in 2017
\citep{lovelace2017}. Extensions of the PCT approach have included
estimation of benefits at the individual level \citep{woodcock2018},
addition of travel to school network \citep{goodman2019}, and improved
modeling of impacts on health, environmental and distributional outcomes
\citep{woodcock2021}. Initially developed just for England, the PCT was
extended to cover all of Wales (for commuter data only) in 2018.

@@

{Another route-based tool is The Cycling Routing Algorithm for Network
Connectivity (CRANC) \citep{gehrke_cycling-focused_2020}, which
evaluates the benefits of new bicycle facilities for the accessibility
of various populations and neighbourhoods.}

@@

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE Tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland (bottom). Source: CRUSE
Tool, publicly available at \href{https://cruse.bike/}{cruse.bike}.}

@@

Following feedback from stakeholders, we added another `non-everyday'
trip purpose in an extension phase: cycle tourism. As outlined in the
CRUSE Extension report {[}reference and link to be added on
publication{]}, this involved developing an inter-county spatial
interaction model to estimate the number of trips between each county
for cycle tourism, with trip attractors including campsites and major
international transport hubs.

@@

The datasets outlined above were used to generate estimates of cycling
currently (the `Baseline' scenario) and under four scenarios of cycling
potential: Near Market, Climate Action Plan, Go Dutch, and Ebike. Each
of these scenarios is described below. Figure~\ref{fig-scenarios}
displays a graph showing the proportion of bicycle trips by distance for
Kildare in each scenario.

@@

\caption{\label{fig-landing}Illustration of the landing page of the
CRUSE Tool for Ireland, showing the default `zoomed out' view of the
national map, with county level results showing the proportion of trips
cycled under the `Go Dutch' scenario (top). The default `zoomed out'
view enables users to click on a county for more detailed results. The
`zoomed in' view (bottom) shows the national network for the `Go Dutch'
scenario in the Limerick area.}

@@

The stakeholder engagement was carried out through workshops with
stakeholders in case study areas, an approach that is documented with
reference to participatory development of digital tools for supporting
energy transitions \citep{hewitt2020}. We held a total of three
workshops over the course of the project and one open-ended live demo
during a 2-day event in Sligo in which we collected feedback from dozens
of people. Other than the live demo, all of the workshops took place
online.

The first two workshops took place in May 2022, during which the tool
was introduced to key staff involved in cycle network planning in
Kildare and Limerick counties. The county-specific approach allowed the
workshop to focus on specific areas and key corridors. The local
transport planners were asked to provide feedback in a semi-structured
interview approach, with questions on the evidence they currently have
access to, how they may be able to use the tool, and requests for new
features. The third and final workshop took place on the
7\textsuperscript{th} December 2023 and included descriptions of recent
changes to the tool, with an aim being to get feedback on the new
recreational and cycle touring evidence, with representatives from
county-level transport planning bodies, Sport Ireland, and Waterways
Ireland.

@@

\includegraphics[width=0.9\textwidth,height=\textheight]{rnet_types.png}
